Causal Discovery on Precarity and Depression

### 2025-02-08 draft_v1/draft_v1.qmd
draft_add_newresult

affiliations:
  - id: 1
    name: \textit{Computational Science Lab, Informatics Institute, University of Amsterdam, PO Box 94323, Amsterdam, 1090GH, the Netherlands}
  - id: 2
    name: \textit{Department of Public and Occupational Health, Amsterdam Public Health Research Institute, Amsterdam UMC, University of Amsterdam, Amsterdam, the Netherland}
  - id: 3
    name: \textit{Institute for Advanced Study, University of Amsterdam, Oude Turfmarkt 147, Amsterdam, 1012GC, the Netherland}   
format: 
  # docx:
  #   number-sections: true
  #   appendix: true
  pdf:
    pdf-engine: pdflatex
    fontfamily: palatino
    toc: true
    number-sections: true
    appendix: true
    colorlinks: true
    keep-tex: true
    template-partials:
      - title.tex
    include-in-header: header.tex
date: today
execute:
  echo: false
  warning: false
bibliography: references_draft.bib
nocite: |
  @*
csl: apa.csl
abstract: |
  \noindent Understanding the causal mechanisms linking precarity factors and depression is critical for developing effective interventions. This study utilizes the HELIUS dataset to explore these relationships using advanced causal discovery methods. By applying algorithms such as FCI, and CCI, and combining traditional Gaussian CI tests with non-parametric approaches like RCoT, we investigate how precarity factors—including employment, social, financial, housing, and relational stress—affect depression, both as a sum score and at the individual symptom level. Our findings reveal that relational stress consistently emerges as a potential causal factor for depression, while symptoms such as sleep disturbances, guilt, and anhedonia are particularly sensitive to external stressors, acting as potential early warning signals or intervention points for prevention. Moreover, the results highlight complexities in the data, including the influence of latent confounders and the challenges of capturing cyclic relationships. Despite some limitations, such as unresolved ambiguities in causal directions and challenges with mixed data distributions, this study demonstrates the utility of causal discovery tools in disentangling the intricate interplay between social and mental health dynamics. By mapping these causal structures into computational models, future research can simulate intervention effects, providing actionable insights to mitigate the impact of precarity on mental health. This study serves as a foundational effort, offering both methodological advancements and practical implications for addressing depression at a population level.
---

@@

Mental health problems in urban areas have been reported to be on the rise. Governments have been attempting to intervene, but the complexity of mental health systems presents significant challenges in planning effective interventions, let alone understanding the underlying mechanisms driving these issues.

Recent research has aimed to identify underlying factors contributing to mental health problems, often referred to as precariousness factors. These factors encompass various aspects of life, such as employment, social connections, financial stability, housing, and cultural dimensions. This comprehensive perspective on precarity helps to highlight how different aspects of life may be interconnected. While this research has advanced our understanding of the complex interplay between precariousness factors, a key question remains unanswered: how do these factors influence mental health? Specifically, the lack of directional information --- knowing what influences what --- limits our ability to identify and prioritize effective intervention targets.

This study aims to investigate the causal relationships between precariousness factors and mental health outcomes, with a specific focus on depression, the most prevalent mental health issue in urban populations. Using causal discovery methods, we explore how different aspects of precarity influence depression and delve deeper into the dynamics at the symptom level. By examining individual depressive symptoms, we aim to identify which symptoms may act as initiators by being particularly sensitive to specific precariousness factors. Through this analysis, our goal is to uncover the causal mechanisms underlying mental health challenges and provide a foundation for developing more effective and targeted interventions.

@@

We use the HELIUS dataset, which captures the diverse population of Amsterdam across various ethnicities and provides comprehensive health and lifestyle data. To operationalize precariousness factors, we draw on the framework outlined in previous research (i.e., Leonie's paper) and select a set of relevant variables.

@@

```{r}
#| layout-ncol: 2
#| label: fig-sum
#| fig-cap: "Resulting graphs of precarity factors and depression sum score using FCI and CCI."
#| fig-subcap: 
#|   - "Using both GaussianCI and RCoT"
#|   - "Using only RCoT"
#| fig-align: center
#| out-width: 100%

knitr::include_graphics("img/both_dot.png")
knitr::include_graphics("img/rcot_dot.png")

@@

<!-- However, due to computational constraints that required limiting the sample size, as well as the challenging distributions of certain symptoms, the statistical power is lower compared to the depression sum score graph. -->
```{r}
#| label: fig-sym
#| fig-cap: "Resulting graphs of precarity factors and individual depression symptoms using FCI and CCI."
#| fig-align: center
#| out-width: 80%

knitr::include_graphics("img/symptom_graph_refined.png")

@@

```{r}
#| layout-nrow: 2
#| label: fig-presum
#| fig-cap: "Resulting graphs of precarity sum score and individual depression symptoms using FCI and CCI."
#| fig-subcap: 
#|   - "FCI"
#|   - "CCI"
#| fig-align: center
#| out-width: 65%

knitr::include_graphics("img/FCI_presum.png")
knitr::include_graphics("img/CCI_presum.png")

@@

@@

```{r}
#| layout-nrow: 2
#| label: fig-pc_sym
#| fig-cap: "Resulting graphs of precarity factors and individual depression symptoms using PC."
#| fig-subcap:
#|   - "Using both GaussianCI and RCoT"
#|   - "Using only RCoT"
#| fig-align: center
#| out-width: 70%

knitr::include_graphics("img/pc_sym_GaussianCIRCoT.png")
knitr::include_graphics("img/pc_sym_onlyRCoT.png")
```

```{r}
#| label: fig-pc_presum
#| fig-cap: "Resulting graphs of precarity sum score and individual depression symptoms using PC."
#| fig-align: center
#| out-width: 70%

@@

## Summarized Stable Edges Proportion {#sec-propmatrix}

```{r}
#| label: fig-sum-mat
#| fig-cap: "Proportions of edge endpoint types for graphs based on depression sum score."
#| fig-align: center
#| out-width: 100%
knitr::include_graphics("img/sumgraph_mat.pdf")
```

```{r}
#| label: fig-sym-mat
#| fig-cap: "Proportions of edge endpoint types for graphs based on individual depression symptoms."
#| fig-align: center
#| out-width: 100%
knitr::include_graphics("img/symgraph_mat.pdf")
```

```{r}
#| label: fig-presum-mat
#| fig-cap: "Proportions of edge endpoint types for graphs based on individual depression symptoms with precarity sum score."
#| fig-align: center
#| out-width: 70%
knitr::include_graphics("img/presumgraph_mat2.pdf")
```

>>>>

affiliations:
  - id: 1
    name: \textit{Computational Science Lab, Informatics Institute, University of Amsterdam, PO Box 94323, Amsterdam, 1090GH, the Netherlands}
  - id: 2
    name: \textit{Department of Public and Occupational Health, Amsterdam Public Health Research Institute, Amsterdam UMC, University of Amsterdam, Amsterdam, the Netherland}
  - id: 3
    name: \textit{Institute for Advanced Study, University of Amsterdam, Oude Turfmarkt 147, Amsterdam, 1012GC, the Netherland}   
format: 
  # docx:
  #   number-sections: true
  #   appendix: true
  pdf:
    pdf-engine: pdflatex
    fontfamily: palatino
    toc: true
    number-sections: true
    appendix: true
    colorlinks: true
    keep-tex: true
    template-partials:
      - title.tex
    include-in-header: header.tex
date: today
execute:
  echo: false
  warning: false
bibliography: references_draft.bib
nocite: |
  @*
csl: apa.csl
abstract: |
  \noindent Understanding the causal mechanisms linking precariousness and depression is critical for developing effective interventions. This study utilizes data from the HELIUS cohort study to explore these relationships using advanced causal discovery methods. By applying algorithms such as FCI, and CCI, and combining traditional Gaussian CI tests with non-parametric approaches like RCoT, we investigate how different indicators of precariousness—including factors related to employment, social relations and relational stress, financial situation, and housing—affect depressed mood, both as a sum score and at the individual symptom level. Our findings reveal that relational stress consistently emerges as a potential causal factor for depressed mood, while symptoms such as sleep disturbances, guilt, and anhedonia are particularly sensitive to external stressors, acting as potential early warning signals or intervention points for prevention. Moreover, the results highlight complexities in the data, including the influence of latent confounders and the challenges of capturing cyclic relationships. Despite some limitations, such as unresolved ambiguities in causal directions and challenges with mixed data distributions, this study demonstrates the utility of causal discovery tools in disentangling the intricate interplay between social and mental health dynamics. By mapping these causal structures into computational models, future research can simulate intervention effects, providing actionable insights to mitigate the impact of precariousness on mental health. This study serves as a foundational effort, offering both methodological advancements and practical implications for addressing depression at a population level.
---

@@

Mental health problems in urban areas have been reported to be on the rise the complexity of mental health systems presents significant challenges in understanding the underlying mechanisms driving these issues, let alone planning effective interventions.

Research has aimed to identify underlying factors contributing to mental health problems. Recent research looked into the association between mental health and indicators of precariousness in various dimensions of life, indicating a high level of uncertainty and instability in people’s lives, such as factors related to employment, social relations, finances, housing, and culture (ref Leonie’s paper). This comprehensive perspective on precariousness helped to highlight how different aspects of life may be interconnected. While this research has advanced our understanding of which indicators of precariousness are related to mental health, a key question remains unanswered: how do these factors influence mental health? Specifically, the lack of directional information — knowing what influences what — limits our ability to identify and prioritize effective intervention targets.

This study aims to investigate the causal relationships between different dimensions of precariousness and mental health outcomes, to with a specific focus on depression, the most prevalent mental health issue. Using causal discovery methods, we explore how different aspects of precariousness influence depression and delve deeper into the dynamics at the symptom level. By examining individual depressive symptoms, we aim to identify which symptoms may act as initiators by being particularly sensitive to precariousness. Through this analysis, our goal is to uncover the causal mechanisms underlying mental health challenges and provide a foundation for developing more effective and targeted interventions.

@@

We use data from the HELIUS study, a multi-ethnic cohort that includes participants of Dutch, Turkish, Moroccan, Surinamese and Ghanaian origin [@snijder2017cohort]. To operationalize precariousness factors, we draw on the framework outlined in previous research (i.e., Leonie's paper) and select a set of relevant variables.

@@

```{r}
#| layout-ncol: 2
#| label: fig-sum
#| fig-cap: "Resulting graph of precarity factors and depression sum score using FCI and proportion of edge endpoint types."
#| fig-subcap: 
#|   - "FCI PAG"
#|   - "Proportion matrix"
#| fig-align: center
#| out-width: 100%

knitr::include_graphics("img/FCI_depsum.png")
knitr::include_graphics("img/depsum_mat_fci.pdf")

@@

<!-- However, due to computational constraints that required limiting the sample size, as well as the challenging distributions of certain symptoms, the statistical power is lower compared to the depression sum score graph. -->
```{r}
#| label: fig-sym
#| layout-nrow: 2
#| fig-cap: "Resulting graph of precarity factors and individual depression symptoms using FCI and proportion of edge endpoint types."
#| fig-subcap: 
#|   - "FCI PAG"
#|   - "Proportion matrix"
#| fig-align: center
#| out-width: 70%

knitr::include_graphics("img/symptom_graph_FCI.png")
knitr::include_graphics("img/symptom_mat_fci.pdf")

@@

```{r}
#| layout-nrow: 2
#| label: fig-presum
#| fig-cap: "Resulting graph of precarity sum score and individual depression symptoms using FCI and proportion of edge endpoint types."
#| fig-subcap: 
#|   - "FCI PAG"
#|   - "Proportion matrix"
#| fig-align: center
#| out-width: 60%

knitr::include_graphics("img/presum_graph_FCI.png")
knitr::include_graphics("img/presum_mat_fci.pdf")
```

@@

## Results from CCI algorithm

```{r}
#| layout-ncol: 2
#| label: fig-sum-cci
#| fig-cap: "Resulting graph of precarity factors and depression sum score using CCI and proportion of edge endpoint types."
#| fig-subcap: 
#|   - "CCI MAAG"
#|   - "Proportion matrix"
#| fig-align: center
#| out-width: 100%

knitr::include_graphics("img/CCI_depsum.png")
knitr::include_graphics("img/depsum_mat_cci.pdf")

```{r}
#| label: fig-sym-cci
#| layout-nrow: 2
#| fig-cap: "Resulting graph of precarity factors and individual depression symptoms using CCI and proportion of edge endpoint types."
#| fig-subcap: 
#|   - "CCI MAAG"
#|   - "Proportion matrix"
#| fig-align: center
#| out-width: 60%

knitr::include_graphics("img/symptom_graph_CCI.png")
knitr::include_graphics("img/symptom_mat_cci.pdf")

```{r}
#| layout-nrow: 2
#| label: fig-presum-cci
#| fig-cap: "Resulting graph of precarity sum score and individual depression symptoms using CCI and proportion of edge endpoint types."
#| fig-subcap: 
#|   - "CCI MAAG"
#|   - "Proportion matrix"
#| fig-align: center
#| out-width: 60%

knitr::include_graphics("img/presum_graph_CCI.png")
knitr::include_graphics("img/presum_mat_cci.pdf")
```

@@

```{r}
#| label: fig-pc_sym
#| fig-cap: "Resulting graphs of precarity factors and individual depression symptoms using PC."
#| fig-align: center
#| out-width: 60%

knitr::include_graphics("img/PC_symptom.png")
```

```{r}
#| label: fig-pc_presum
#| fig-cap: "Resulting graphs of precarity sum score and individual depression symptoms using PC."
#| fig-align: center
#| out-width: 60%

@@

<!-- ## Summarized Stable Edges Proportion {#sec-propmatrix} -->

<!-- ```{r} -->
<!-- #| label: fig-sum-mat -->
<!-- #| fig-cap: "Proportions of edge endpoint types for graphs based on depression sum score." -->
<!-- #| fig-align: center -->
<!-- #| out-width: 100% -->
<!-- knitr::include_graphics("img/sumgraph_mat.pdf") -->
<!-- knitr::include_graphics("img/sumgraph_mat.pdf") -->

<!-- ``` -->

<!-- ```{r} -->
<!-- #| label: fig-sym-mat -->
<!-- #| fig-cap: "Proportions of edge endpoint types for graphs based on individual depression symptoms." -->
<!-- #| fig-align: center -->
<!-- #| out-width: 100% -->
<!-- knitr::include_graphics("img/symgraph_mat.pdf") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #| label: fig-presum-mat -->
<!-- #| fig-cap: "Proportions of edge endpoint types for graphs based on individual depression symptoms with precarity sum score." -->
<!-- #| fig-align: center -->
<!-- #| out-width: 70% -->
<!-- knitr::include_graphics("img/presumgraph_mat2.pdf") -->
<!-- ``` -->

### 2025-02-08 draft_v1/draft_v1.tex
draft_add_newresult

\date{2025-01-16}
\begin{document}
\maketitle
\begin{abstract}
\noindent Understanding the causal mechanisms linking precarity factors
and depression is critical for developing effective interventions. This
study utilizes the HELIUS dataset to explore these relationships using
advanced causal discovery methods. By applying algorithms such as FCI,
and CCI, and combining traditional Gaussian CI tests with non-parametric
approaches like RCoT, we investigate how precarity factors---including
employment, social, financial, housing, and relational stress---affect
depression, both as a sum score and at the individual symptom level. Our
findings reveal that relational stress consistently emerges as a
potential causal factor for depression, while symptoms such as sleep
disturbances, guilt, and anhedonia are particularly sensitive to
external stressors, acting as potential early warning signals or
intervention points for prevention. Moreover, the results highlight
complexities in the data, including the influence of latent confounders
and the challenges of capturing cyclic relationships. Despite some
limitations, such as unresolved ambiguities in causal directions and
challenges with mixed data distributions, this study demonstrates the
utility of causal discovery tools in disentangling the intricate
interplay between social and mental health dynamics. By mapping these
causal structures into computational models, future research can
simulate intervention effects, providing actionable insights to mitigate
the impact of precarity on mental health. This study serves as a
foundational effort, offering both methodological advancements and
practical implications for addressing depression at a population level.
\end{abstract}

@@

Mental health problems in urban areas have been reported to be on the
rise. Governments have been attempting to intervene, but the complexity
of mental health systems presents significant challenges in planning
effective interventions, let alone understanding the underlying
mechanisms driving these issues.

Recent research has aimed to identify underlying factors contributing to
mental health problems, often referred to as precariousness factors.
These factors encompass various aspects of life, such as employment,
social connections, financial stability, housing, and cultural
dimensions. This comprehensive perspective on precarity helps to
highlight how different aspects of life may be interconnected. While
this research has advanced our understanding of the complex interplay
between precariousness factors, a key question remains unanswered: how
do these factors influence mental health? Specifically, the lack of
directional information --- knowing what influences what --- limits our
ability to identify and prioritize effective intervention targets.

This study aims to investigate the causal relationships between
precariousness factors and mental health outcomes, with a specific focus
on depression, the most prevalent mental health issue in urban
populations. Using causal discovery methods, we explore how different
aspects of precarity influence depression and delve deeper into the
dynamics at the symptom level. By examining individual depressive
symptoms, we aim to identify which symptoms may act as initiators by
being particularly sensitive to specific precariousness factors. Through
this analysis, our goal is to uncover the causal mechanisms underlying
mental health challenges and provide a foundation for developing more
effective and targeted interventions.

@@

We use the HELIUS dataset, which captures the diverse population of
Amsterdam across various ethnicities and provides comprehensive health
and lifestyle data. To operationalize precariousness factors, we draw on
the framework outlined in previous research (i.e., Leonie's paper) and
select a set of relevant variables.

@@

This setup yields 16 combinations (2 significance levels × 4 thresholds
× 2 CI tests), applied across three algorithms. Each combination is
repeated for 100 bootstrap samples, yielding a total of 1,600 graphs per
algorithm. For analyses involving individual symptom variables, we
simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing
the number of bootstrap samples to 30, easing computational demands. To
further enhance efficiency, we fix the skeleton of edges among symptom
variables based on the common structure estimated across all algorithms.
This approach minimizes spurious edges in the symptom network and
significantly reduces the computational time needed for skeleton
estimation. This fixed structure also aligns well with commonly reported
skeleton structures in the literature, ensuring consistency with
existing findings (\textbf{cite the comp\_model paper}).\footnote{We
  fixed the skeleton of the symptom network using a common structure
  derived from the PC, FCI, and CCI algorithms with an alpha level of
  0.001. This approach preserved the complexity of symptom interactions,
  enabling unrestricted estimation of causal directions both among
  symptoms (\emph{symptom --- symptom} interactions) and between
  symptoms and precarity factors.} Lastly, to summarize the results, we
identify the most frequently occurring edge endpoints across different
experimental setups.\footnote{The detailed proportion of each edge
  endpoint occurrence is provided in Section~\ref{sec-propmatrix}.} This
ensures that only stable and consistent edges are retained, providing a
clearer and more reliable understanding of the relationships between
precarity factors and depression. See Figure~\ref{fig-workflow} for an
overview of the analysis workflow.

@@

\includegraphics[width=1\textwidth,height=\textheight]{img/both_dot.png}

@@

\subcaption{\label{fig-sum-1}Using both GaussianCI and RCoT}

@@

\includegraphics[width=1\textwidth,height=\textheight]{img/rcot_dot.png}

@@

\subcaption{\label{fig-sum-2}Using only RCoT}

@@

\caption{\label{fig-sum}Resulting graphs of precarity factors and
depression sum score using FCI and CCI.}

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{img/symptom_graph_refined.png}

@@

\caption{\label{fig-sym}Resulting graphs of precarity factors and
individual depression symptoms using FCI and CCI.}

@@

Compared to Figure~\ref{fig-sym}, this analysis reveals greater
disagreement between FCI and CCI regarding edge directionality, as
evidenced by more dashed edges. FCI tends to estimate more undirected
edges (circles), while CCI more frequently assigns directed arrows
between symptom-symptom relationships. For a detailed breakdown, see
Section~\ref{sec-propmatrix}, which shows the proportion of symbols
produced by each algorithm. When examining overall precarity, we observe
that more symptoms are now connected to the precarity score
(\emph{Precarity}) compared to the individual precarity factor analysis
in Figure~\ref{fig-sym}. In addition to \emph{glt}, \emph{slp},
\emph{anh}, and \emph{sui}, which are previously shown to be connected
to individual precarity factors, \emph{mot}, \emph{app}, and \emph{dep}
are now identified as having causal relationships with overall
precarity. Furthermore, when using the Gaussian CI test, con is also
shown to be causally related to overall precarity. Notably, all these
connections are directed towards precarity (o→), indicating that overall
precarity is not the cause of symptoms. The exception is the
bidirectional edge between \emph{sui} \textless-\textgreater{}
\emph{Precarity}, which suggests a connection via latent confounders.
FCI predominantly estimates directed arrows (→) for symptom-precarity
relationships, suggesting that symptoms are driving precarity. With
fewer variables (reduced from 14 to 9), this approach appears to have
slightly more statistical power, enabling algorithms to estimate
relationships with greater confidence compared to the previous analysis
involving individual symptoms and individual precarity factors.

@@

\includegraphics[width=0.65\textwidth,height=\textheight]{img/FCI_presum.png}

@@

\subcaption{\label{fig-presum-1}FCI}

@@

\includegraphics[width=0.65\textwidth,height=\textheight]{img/CCI_presum.png}

@@

\subcaption{\label{fig-presum-2}CCI}

@@

\caption{\label{fig-presum}Resulting graphs of precarity sum score and
individual depression symptoms using FCI and CCI.}

@@

@@

@@

\includegraphics[width=0.7\textwidth,height=\textheight]{img/PC_symptom.png}

@@

\includegraphics[width=0.7\textwidth,height=\textheight]{img/PC_presum.png}

@@

\subsection{Summarized Stable Edges Proportion}\label{sec-propmatrix}

\includegraphics[width=1\textwidth,height=\textheight]{img/sumgraph_mat.pdf}

\caption{\label{fig-sum-mat}Proportions of edge endpoint types for
graphs based on depression sum score.}

\includegraphics[width=1\textwidth,height=\textheight]{img/symgraph_mat.pdf}

\caption{\label{fig-sym-mat}Proportions of edge endpoint types for
graphs based on individual depression symptoms.}

\includegraphics[width=0.7\textwidth,height=\textheight]{img/presumgraph_mat2.pdf}

\caption{\label{fig-presum-mat}Proportions of edge endpoint types for
graphs based on individual depression symptoms with precarity sum
score.}

>>>>

\date{2025-02-08}
\begin{document}
\maketitle
\begin{abstract}
\noindent Understanding the causal mechanisms linking precariousness and
depression is critical for developing effective interventions. This
study utilizes data from the HELIUS cohort study to explore these
relationships using advanced causal discovery methods. By applying
algorithms such as FCI, and CCI, and combining traditional Gaussian CI
tests with non-parametric approaches like RCoT, we investigate how
different indicators of precariousness---including factors related to
employment, social relations and relational stress, financial situation,
and housing---affect depressed mood, both as a sum score and at the
individual symptom level. Our findings reveal that relational stress
consistently emerges as a potential causal factor for depressed mood,
while symptoms such as sleep disturbances, guilt, and anhedonia are
particularly sensitive to external stressors, acting as potential early
warning signals or intervention points for prevention. Moreover, the
results highlight complexities in the data, including the influence of
latent confounders and the challenges of capturing cyclic relationships.
Despite some limitations, such as unresolved ambiguities in causal
directions and challenges with mixed data distributions, this study
demonstrates the utility of causal discovery tools in disentangling the
intricate interplay between social and mental health dynamics. By
mapping these causal structures into computational models, future
research can simulate intervention effects, providing actionable
insights to mitigate the impact of precariousness on mental health. This
study serves as a foundational effort, offering both methodological
advancements and practical implications for addressing depression at a
population level.
\end{abstract}

@@

Mental health problems in urban areas have been reported to be on the
rise the complexity of mental health systems presents significant
challenges in understanding the underlying mechanisms driving these
issues, let alone planning effective interventions.

Research has aimed to identify underlying factors contributing to mental
health problems. Recent research looked into the association between
mental health and indicators of precariousness in various dimensions of
life, indicating a high level of uncertainty and instability in people's
lives, such as factors related to employment, social relations,
finances, housing, and culture (ref Leonie's paper). This comprehensive
perspective on precariousness helped to highlight how different aspects
of life may be interconnected. While this research has advanced our
understanding of which indicators of precariousness are related to
mental health, a key question remains unanswered: how do these factors
influence mental health? Specifically, the lack of directional
information --- knowing what influences what --- limits our ability to
identify and prioritize effective intervention targets.

This study aims to investigate the causal relationships between
different dimensions of precariousness and mental health outcomes, to
with a specific focus on depression, the most prevalent mental health
issue. Using causal discovery methods, we explore how different aspects
of precariousness influence depression and delve deeper into the
dynamics at the symptom level. By examining individual depressive
symptoms, we aim to identify which symptoms may act as initiators by
being particularly sensitive to precariousness. Through this analysis,
our goal is to uncover the causal mechanisms underlying mental health
challenges and provide a foundation for developing more effective and
targeted interventions.

@@

We use data from the HELIUS study, a multi-ethnic cohort that includes
participants of Dutch, Turkish, Moroccan, Surinamese and Ghanaian origin
(Snijder et al., 2017). To operationalize precariousness factors, we
draw on the framework outlined in previous research (i.e., Leonie's
paper) and select a set of relevant variables.

@@

This setup yields 16 combinations (2 significance levels × 4 thresholds
× 2 CI tests), applied across three algorithms. Each combination is
repeated for 100 bootstrap samples, yielding a total of 1,600 graphs per
algorithm. For analyses involving individual symptom variables, we
simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing
the number of bootstrap samples to 30, easing computational demands. To
further enhance efficiency, we fix the skeleton of edges among symptom
variables based on the common structure estimated across all algorithms.
This approach minimizes spurious edges in the symptom network and
significantly reduces the computational time needed for skeleton
estimation. This fixed structure also aligns well with commonly reported
skeleton structures in the literature, ensuring consistency with
existing findings (\textbf{cite the comp\_model paper}).\footnote{We
  fixed the skeleton of the symptom network using a common structure
  derived from the PC, FCI, and CCI algorithms with an alpha level of
  0.001. This approach preserved the complexity of symptom interactions,
  enabling unrestricted estimation of causal directions both among
  symptoms (\emph{symptom --- symptom} interactions) and between
  symptoms and precarity factors.} Lastly, to summarize the results, we
identify the most frequently occurring edge endpoints across different
experimental setups.\footnote{The detailed proportion of each edge
  endpoint occurrence is provided in \textbf{?@sec-propmatrix}.} This
ensures that only stable and consistent edges are retained, providing a
clearer and more reliable understanding of the relationships between
precarity factors and depression. See Figure~\ref{fig-workflow} for an
overview of the analysis workflow.

@@

\includegraphics[width=1\textwidth,height=\textheight]{img/FCI_depsum.png}

@@

\subcaption{\label{fig-sum-1}FCI PAG}

@@

\includegraphics[width=1\textwidth,height=\textheight]{img/depsum_mat_fci.pdf}

@@

\subcaption{\label{fig-sum-2}Proportion matrix}

@@

\caption{\label{fig-sum}Resulting graph of precarity factors and
depression sum score using FCI and proportion of edge endpoint types.}

@@

\includegraphics[width=0.7\textwidth,height=\textheight]{img/symptom_graph_FCI.png}

@@

\subcaption{\label{fig-sym-1}FCI PAG}

\includegraphics[width=0.7\textwidth,height=\textheight]{img/symptom_mat_fci.pdf}

\subcaption{\label{fig-sym-2}Proportion matrix}

\caption{\label{fig-sym}Resulting graph of precarity factors and
individual depression symptoms using FCI and proportion of edge endpoint
types.}

@@

Compared to Figure~\ref{fig-sym}, this analysis reveals greater
disagreement between FCI and CCI regarding edge directionality, as
evidenced by more dashed edges. FCI tends to estimate more undirected
edges (circles), while CCI more frequently assigns directed arrows
between symptom-symptom relationships. For a detailed breakdown, see
\textbf{?@sec-propmatrix}, which shows the proportion of symbols
produced by each algorithm. When examining overall precarity, we observe
that more symptoms are now connected to the precarity score
(\emph{Precarity}) compared to the individual precarity factor analysis
in Figure~\ref{fig-sym}. In addition to \emph{glt}, \emph{slp},
\emph{anh}, and \emph{sui}, which are previously shown to be connected
to individual precarity factors, \emph{mot}, \emph{app}, and \emph{dep}
are now identified as having causal relationships with overall
precarity. Furthermore, when using the Gaussian CI test, con is also
shown to be causally related to overall precarity. Notably, all these
connections are directed towards precarity (o→), indicating that overall
precarity is not the cause of symptoms. The exception is the
bidirectional edge between \emph{sui} \textless-\textgreater{}
\emph{Precarity}, which suggests a connection via latent confounders.
FCI predominantly estimates directed arrows (→) for symptom-precarity
relationships, suggesting that symptoms are driving precarity. With
fewer variables (reduced from 14 to 9), this approach appears to have
slightly more statistical power, enabling algorithms to estimate
relationships with greater confidence compared to the previous analysis
involving individual symptoms and individual precarity factors.

@@

\includegraphics[width=0.6\textwidth,height=\textheight]{img/presum_graph_FCI.png}

@@

\subcaption{\label{fig-presum-1}FCI PAG}

@@

\includegraphics[width=0.6\textwidth,height=\textheight]{img/presum_mat_fci.pdf}

@@

\subcaption{\label{fig-presum-2}Proportion matrix}

@@

\caption{\label{fig-presum}Resulting graph of precarity sum score and
individual depression symptoms using FCI and proportion of edge endpoint
types.}

@@

\bibitem[\citeproctext]{ref-snijder2017cohort}
Snijder, M. B., Galenkamp, H., Prins, M., Derks, E. M., Peters, R. J.,
Zwinderman, A. H., \& Stronks, K. (2017). Cohort profile: The healthy
life in an urban setting (HELIUS) study in amsterdam, the netherlands.
\emph{BMJ Open}, \emph{7}(12), e017873.

@@

\subsection{Results from CCI
algorithm}\label{results-from-cci-algorithm}

\includegraphics[width=1\textwidth,height=\textheight]{img/CCI_depsum.png}

\subcaption{\label{fig-sum-cci-1}CCI MAAG}

\includegraphics[width=1\textwidth,height=\textheight]{img/depsum_mat_cci.pdf}

\subcaption{\label{fig-sum-cci-2}Proportion matrix}

\caption{\label{fig-sum-cci}Resulting graph of precarity factors and
depression sum score using CCI and proportion of edge endpoint types.}

\includegraphics[width=0.6\textwidth,height=\textheight]{img/symptom_graph_CCI.png}

\subcaption{\label{fig-sym-cci-1}CCI MAAG}

\includegraphics[width=0.6\textwidth,height=\textheight]{img/symptom_mat_cci.pdf}

\subcaption{\label{fig-sym-cci-2}Proportion matrix}

\caption{\label{fig-sym-cci}Resulting graph of precarity factors and
individual depression symptoms using CCI and proportion of edge endpoint
types.}

\includegraphics[width=0.6\textwidth,height=\textheight]{img/presum_graph_CCI.png}

\subcaption{\label{fig-presum-cci-1}CCI MAAG}

\includegraphics[width=0.6\textwidth,height=\textheight]{img/presum_mat_cci.pdf}

\subcaption{\label{fig-presum-cci-2}Proportion matrix}

\caption{\label{fig-presum-cci}Resulting graph of precarity sum score
and individual depression symptoms using CCI and proportion of edge
endpoint types.}

@@

\includegraphics[width=0.6\textwidth,height=\textheight]{img/PC_symptom.png}

@@

\includegraphics[width=0.6\textwidth,height=\textheight]{img/PC_presum.png}

@@
